---
title: "Introductory Categorical Data Analysis in R"
author: "Clay Ford"
date: "Fall 2016"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Categorical Data

In categorical data analysis, our response of interest (or dependent variable) is categorical.   
\ 

The measurement scale of categorical data is a set of categories.   
\ 

Categories can be _nominal_ or _ordered_.  
\ 

Nominal example: north, south, east, west  
Ordinal example: mild, medium, hot, very hot  
\ 

Whether your data are nominal or ordinal determines what kinds of statistical analyses you should use.


## Nominal/Ordinal Scale Distinction

Order of categories is irrelevant for nominal data and does not affect statistical analysis.   
\ 

Order of categories is important for ordinal data and does affect the statistical analysis.   
\ 

Methods designed for ordinal data cannot be used with nominal data. The reverse is not true, but the analysis is not optimal since it does not utilize the ordering information. 

## Categorical data summaries and analyses

We typically tabulate category membership and calculate proportions.   
\ 

Examples of analyses:

- compare proportions between two categories
- examine the ratio of proportions or odds between two categories
- look for an assocation between two categorical variables
- model category proportions as a function of other variables
- model category counts as a function of other variables

The big idea of categorical data analysis is to find out if the distribution of category membership is explained by or associated with other variables. 


## Categorical data analysis in R

Three steps:

1. get data into R
2. manipulate it (if necessary) so it's suitable for analysis and/or visualization
3. run the analysis and/or create some graphs

Today we'll look at some strategies for each step that are useful for categorical data analysis.

## Example 1

Relationship between aspirin use and myocardial infarction (heart attacks) by the Physicians' Health Study Research Group at Harvard Medical School (_NEJM_, 318:262-264, 1988). Five year randomized study testing whether regular intake of aspirin reduces mortality from cardiovascular disease.

```{r echo=FALSE}
aspirin <- matrix(data = c(189, 104, 10845, 10933), 
                  ncol=2, 
                  dimnames = list("group" = c("Placebo","Aspirin"),
                                  "MI" = c("Yes","No")))
```


```{r}
aspirin
```

Are these two categorical variables associated?
 
## Some analysis options

1. Compare proportion of MI for placebo group versus proportion of MI for aspirin group. (Difference of Proportions)
2. Examine ratio of MI proportions for both groups. (Risk Ratio or Relative Risk)
3. Examine ratio of the estimated odds of MI for both groups. (Odds Ratio)
4. Hypothesis test of independence between two variables (Chi-Square Test)

## Getting the data into R "by hand"

We can enter this manually if we like.

```{r echo=TRUE}
aspirin <- matrix(data = c(189, 104, 10845, 10933), 
                  ncol = 2, 
                  dimnames = list("group" = 
                                    c("Placebo","Aspirin"),
                                  "MI" = 
                                    c("Yes","No")))
```

We now have a _contingency table_ in the form of a matrix ready for analysis.

## Getting the data into R by import

Data is often stored in txt, csv or xlsx formats, or stored in another stats program. There are several ways to import such data into R.

- `read.table`
- `read.csv`
- `read_csv` in the `readr` package
- `read_excel` in the `readxl` package
- `read_spss`, `read_sas`, and `read_stata` in the `haven` package

Importing data in this manner creates a _data frame_. To form a contingency table from variables stored in a data frame, we can use the `table` or `xtabs` functions.

## The `table` function

`table` forms a contingency table from vectors. Say we have a data frame called `dat` with columns called `X` and `Y`.  
```{r echo=FALSE}
set.seed(3)
dat <- data.frame(X = sample(c("Ctrl","Trt"),10, replace = T), Y = sample(c("Yes","No"),10, replace = T))
```

```{r echo=TRUE}
dat$X
dat$Y
```



`table(dat$X, dat$Y)` creates a 2 x 2 table with counts of `X` on the rows and counts of `Y` on the columns.

## `table` example

```{r echo=TRUE}
table(dat$X, dat$Y)
```

This works too:
```{r results='hide', echo=TRUE}
with(dat, table(X, Y))

```

## The `xtabs` function

The `xtabs` function can create a table in two different ways: 

1. from a data frame where there is one observation per record
2. from a data frame where each row represents total counts

Syntax:    

- `xtabs( ~ X + Y, data = dat)` - one obs per record   
- `xtabs(Freq ~ X + Y, data = dat)` - each row has total counts stored in column called `Freq`

## `xtabs` example 1

This returns the same thing as `table(dat$X, dat$Y)` except with rows and columns labeled:

```{r echo=TRUE}
xtabs(~ X + Y, data = dat)
```

## `xtabs` example 2

Say our data looks like this:

```{r echo=FALSE}
dat2 <- as.data.frame(table(dat$X, dat$Y))
names(dat2)[1:2] <- c("X","Y")
```

```{r echo=TRUE}
dat2
```

## `xtabs` example 2

We can use `xtabs` to create a table like so:

```{r echo=TRUE}
xtabs(Freq ~ X + Y, data = dat2)
```


## working with tables

We sometimes want to work with tables to find proportions or marginal totals. Good functions to know include:

- `prop.table` (cell proportions)
- `margin.table` (get table margins)
- `addmargins` (add table margins)

## `prop.table` example

Specify `margin = 1` for row-wise proportions, `margin = 2` for column-wise proportions.

```{r echo=TRUE}
prop.table(aspirin, margin = 1)
```

## `margin.table` example

Specify `margin = 1` for row-wise margins, `margin = 2` for column-wise margins.

```{r echo=TRUE}
margin.table(aspirin, margin = 1)
margin.table(aspirin, margin = 2)

```

## `addmargins` example

The basic usage is to add row- and column-wise totals as well as the table total. 

```{r echo=TRUE}
addmargins(aspirin)
```

See `?addmargins` for much advanced usage.

## Analysis: Chi-square test

The chi-square test of independence tests the null hypothesis that two categorical variables are not associated with one another. We can use the `chisq.test` for this.

```{r echo=TRUE}
chisq.test(aspirin)
```

A low p-value rejects the null of no association. Notice it doesn't tell you the strength of association or the direction. A Chi-square test is rarely sufficient for answering all the questions you'll have about your data.

## Analysis: comparing proportions

We can use the `prop.test` function to compare proportions. It can take a two-dimensional table (or matrix) with 2 columns, giving the counts of successes and failures, respectively.

```{r echo=TRUE}
prop.test(aspirin)
```

## Analysis: risk ratios

A difference in proportions may have greater importance when both proportions are near 0 or 1. The risk ratio is a useful descriptive measure in this case.   
\ 

Example, both differ by `0.009`:
```{r echo=TRUE}
0.410 / 0.401
0.010 / 0.001
```

When risk ratio = 1, we conclude there is no difference in proportions.

## Analysis: risk ratios

Risk ratio for aspirin data reveals the proportion of MI for placebo group was about 82% higher:

```{r echo=TRUE}
prop.table(aspirin, margin = 1)[1,1] / 
  prop.table(aspirin, margin = 1)[2,1]

```

We'll use the `epitools` package for easier calculation of risk ratios and to obtain confidence intervals.

## Analysis: odds ratios

The odds are related to probability as follows:

$$ odds = \frac{p}{1 - p} $$

For example, when $p = 0.75$:

$$ 3 = \frac{0.75}{0.25} $$

We expect to observe 3 successes for every 1 failure. The ratio of two odds is called an _odds ratio_. An odds ratio of 1 implies there is no difference between the odds.

## Analysis: odds ratios

A quick formula for estimating an odds ratio in a 2 x 2 table is

$$\hat{\theta} = \frac{n_{11} n_{22}}{n_{12} n_{21}} $$

For our aspirin data, the odds of MI in the placebo group is about 83% higher than the odds of MI in the aspirin group:

```{r echo=TRUE}
(aspirin[1,1] * aspirin[2,2]) / 
  (aspirin[1,2] * aspirin[2,1])
```

Again, we'll use the `epitools` package for easier calculation of odds ratios and to obtain confidence intervals. Let's go to R!

## Example 2

Eight studies in China about smoking and lung cancer. (_Intern. J. Epidemiol._, 21: 197-201, 1992) We want to investigate the relationship between smoking and lung cancer while controlling for city.

```{r}
lung.cancer <- array(data = c(126,35,100,61,
                              908,497,688,807,
                              913,336,747,598,
                              235,58,172,121,
                              402,121,308,215,
                              182,72,156,98,
                              60,11,99,43,
                              104,21,89,36),
                     dim = c(2,2,8),
                     dimnames = list("smoking" = c("smokers","nonsmokers"),
                                  "lung.cancer" = c("yes","no"),
                                  "city" = c("Beijing","Shanghai","Shenyang",
                                             "Nanjing","Harbin","Zhengzhou",
                                             "Taiyuan","Nanchang")))
# Table 3.3
ftable(lung.cancer, row.vars = c("city","smoking"))
```

## Some analysis options

1. Test the null hypothesis that smoking and lung cancer are conditionally independent given city, that is the odds ratio = 1 for each _partial table_ (Cochran-Mantel-Haenszel Test)
2. Test the null hypothesis that the odds ratio between smoking and lung cancer is the same at each level of city. (Breslow-Day Test)
3. Estimate a _common odds ratio_ is association seems stable across the partial tables.

## Getting data into R

We demonstrate how to get this data into R in the R script that accompanies this workshop. The code doesn't fit comfortably onto slides.    
\ 


The manual method involves the `array` function that allows us to specify a third dimension, or third layer, of data.   
\ 

Once again we can use the `table` and `xtabs` functions to create our contingency tables if we import our data into a data frame.

## The `table` function for 3-way tables

```{r echo=FALSE}
set.seed(1)
dat3 <- data.frame(X = sample(c("Ctrl","Trt"),10, replace = T), Y = sample(c("Yes","No"),10, replace = T),
                  Z = sample(c("Male","Female"), 10, replace = T))
```

The first vector in the `table` function forms the rows, the second the columns, and each subsequent vector a layer.

```{r echo=TRUE}
table(dat3$X, dat3$Y, dat3$Z)
```

## The `xtabs` function for 3-way tables



```{r echo=TRUE}
# one row per obs
xtabs( ~ X + Y + Z, data = dat3) 
```

## The `xtabs` function for 3-way tables


```{r}
dat4 <- as.data.frame(table(dat3$X, dat3$Y, dat3$Z))
names(dat4)[1:3] <- c("X","Y","Z")
```

```{r echo=TRUE}
# one row per cell Freq
xtabs(Freq ~ X + Y + Z, data = dat4) 
```


## Example 3

Logistic regression

## Topics we haven't covered


## References

Agresti, A. _An Introduction to Categorical Data Analysis_. (1996) - **The source for much of this workshop**.
