---
title: "Introductory Categorical Data Analysis in R"
author: "Clay Ford"
date: "Fall 2016"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## workshop outline

Topics on tap:

- Analysis of two-way contingency tables
- Analysis of three-way contingency tables
- Logistic regression


## Categorical Data

In categorical data analysis, our response of interest (or dependent variable) is categorical.   
\ 

The measurement scale of categorical data is a set of categories.   
\ 

Categories can be _nominal_ or _ordered_.  

- Nominal example: Thai, Mexican, Italian, Chinese  
- Ordinal example: mild, medium, hot, very hot  

Whether your data are nominal or ordinal determines what kinds of statistical analyses you should use. Today we'll mainly focus on nominal data with two categories.

## Categorical data summaries and analyses

We typically tabulate category membership and calculate proportions.   
\ 

Examples of analyses:

- compare proportions between two categories
- examine the ratio of proportions or odds between two categories
- look for an assocation between two categorical variables
- model category proportions as a function of other variables
- model category counts as a function of other variables

The big idea of categorical data analysis is to find out if the distribution of category membership is explained by or associated with other variables. 


## Categorical data analysis in R

Three steps:

1. get data into R
2. manipulate it (if necessary) so it's suitable for analysis and/or visualization
3. run the analysis and/or create some graphs

Today we'll look at some strategies for each step that are useful for categorical data analysis.

## Example 1

Relationship between aspirin use and myocardial infarction (heart attacks) by the Physicians' Health Study Research Group at Harvard Medical School (_NEJM_, 318:262-264, 1988). Five year randomized study testing whether regular intake of aspirin reduces mortality from cardiovascular disease.

```{r echo=FALSE}
aspirin <- matrix(data = c(189, 104, 10845, 10933), 
                  ncol=2, 
                  dimnames = list("group" = c("Placebo","Aspirin"),
                                  "MI" = c("Yes","No")))
```


```{r}
aspirin
```

Are these two categorical variables associated?
 
## Some analysis options

1. Hypothesis test of independence between two variables (Chi-Square Test)
2. Compare proportion of MI for placebo group versus proportion of MI for aspirin group. (Difference of Proportions)
3. Examine ratio of MI proportions for both groups. (Risk Ratio or Relative Risk)
4. Examine ratio of the estimated odds of MI for both groups. (Odds Ratio)

## Getting the data into R "by hand"

We can enter this manually if we like.

```{r echo=TRUE}
aspirin <- matrix(data = c(189, 104, 10845, 10933), 
                  ncol = 2, 
                  dimnames = list("group" = 
                                    c("Placebo","Aspirin"),
                                  "MI" = 
                                    c("Yes","No")))
```

We now have a _contingency table_ in the form of a matrix ready for analysis.

## Our matrix

```{r echo=TRUE}
aspirin
```

Notice the dimensions are labeled and the rows are in the order we specified.

## Getting the data into R by import

Data is often stored in txt, csv or xlsx formats, or stored in another stats program. There are several ways to import such data into R.

- `read.table`
- `read.csv`
- `read_csv` in the `readr` package
- `read_excel` in the `readxl` package
- `read_spss`, `read_sas`, and `read_stata` in the `haven` package

Importing data in this manner creates a _data frame_. To form a contingency table from variables stored in a data frame, we can use the `table` or `xtabs` functions.

## Example of `read.csv` for importing data

If your csv file has a header, and the csv file is in your working directory, then we simply need to give read.csv the name of the csv file. 

```{r echo=TRUE}
aspirin2 <- read.csv("aspirin.csv")
head(aspirin2)
```



## The `table` function

`table` forms a contingency table from vectors. The first vector becomes the rows and the second vector the columns. Notice the rows and columns are placed in alphabetical order:

```{r echo=TRUE}
table(aspirin2$group, aspirin2$MI)
```

This works too:
```{r results='hide', echo=TRUE}
with(aspirin2, table(group, MI))

```


## The `xtabs` function

The `xtabs` function can create a table in two different ways: 

1. from a data frame where there is one observation per record
2. from a data frame where each row represents total counts

Syntax:    

- `xtabs( ~ X + Y, data = dat)` - one obs per record   
- `xtabs(Freq ~ X + Y, data = dat)` - each row has total counts stored in column called `Freq`

## `xtabs` example 1

This returns the same thing as `table(aspirin2$group, aspirin2$MI)` except with rows and columns labeled:

```{r echo=TRUE}
xtabs(~ group + MI, data = aspirin2)
```

## `xtabs` example 2

Say our data frame looks like this:

```{r echo=FALSE}
aspirin3 <- as.data.frame(table(aspirin2$group, aspirin2$MI))
names(aspirin3)[1:2] <- c("group","MI")
```

```{r echo=TRUE}
aspirin3
```

## `xtabs` example 2

We can use `xtabs` to create a table like so:

```{r echo=TRUE}
xtabs(Freq ~ group + MI, data = aspirin3)
```


## working with 2-way tables

We sometimes want to work with tables to find proportions or marginal totals. Good functions to know include:

- `prop.table` (cell proportions)
- `margin.table` (get table margins)
- `addmargins` (add table margins)

## `prop.table` example

Specify `margin = 1` for row-wise proportions, `margin = 2` for column-wise proportions. Not specifying `margin` returns cell proportions summing to 1.

```{r echo=TRUE}
prop.table(aspirin, margin = 1)
```

## `margin.table` example

Specify `margin = 1` for row-wise margins, `margin = 2` for column-wise margins. Not specifying `margin` sums all cell totals.

```{r echo=TRUE}
margin.table(aspirin, margin = 1)
margin.table(aspirin, margin = 2)

```

## `addmargins` example

The basic usage is to add row- and column-wise totals as well as the table total. Specifying `margin=1` or `margin=2` shows only row or column totals, respectively.

```{r echo=TRUE}
addmargins(aspirin)
```

See `?addmargins` for advanced usage.

## Analysis: Chi-square test

The chi-square test of independence tests the null hypothesis that two categorical variables are not associated with one another. We can use the `chisq.test` function for this.

```{r echo=TRUE}
chisq.test(aspirin)
```

A low p-value rejects the null of no association. Notice it doesn't tell you the strength of association or the direction. A Chi-square test is rarely sufficient for answering all the questions you'll have about your data.

## Analysis: comparing proportions

Use the `prop.test` function to compare proportions of "successes". It can take a 2 x 2 table (or matrix) with counts of successes and failures, respectively, in the columns.

```{r echo=TRUE}
prop.test(aspirin)
```

## A drawback of difference in proportions

Using the absolute difference between proportions to compare two groups can be misleading. Example:

- The difference between 0.410 and 0.401 is 0.009.
- The difference between 0.010 and 0.001 is also 0.009.

However the latter difference seems more important if we compare ratios:

- 0.410/0.401 = 1.02244
- 0.010/0.001 = 10

0.010 is 10 times larger than 0.001. A difference in proportions may have greater importance when both proportions are near 0 or 1.

## Analysis: risk ratios

The _risk ratio_ (or _relative risk_) is simply the proportion of successes in one group divided the proportion of successes in the other group. A proportion of 1 implies no difference between the groups.    
\ 

Risk ratio for aspirin data reveals the proportion of MI for placebo group was about 82% higher:

```{r echo=TRUE}
prop.table(aspirin, margin = 1)[1,1] / 
  prop.table(aspirin, margin = 1)[2,1]

```

We'll use the `epitools` package for easier calculation of risk ratios and to obtain confidence intervals.

## odds

We can also compare odds of success in one group versus odds of success in another group.   
\ 

The odds are related to probability as follows:

$$ odds = \frac{p}{1 - p} $$

- When $p = 0.75$, the odds are $\frac{0.75}{0.25} =3$. We expect to observe 3 successes for every 1 failure. 
- When $p = 0.25$, the odds are $\frac{0.25}{0.75} \approx \frac{1}{3}$. We expect to observe about 1 success for every 3 failures. 

## odds ratios

When using odds to compare groups, we typically take the ratio of the two odds.   
\ 

The ratio of two odds is called an _odds ratio_. An odds ratio of 1 implies there is no difference between the odds of success in the two groups.    
\ 

A quick formula for calculating the odds ratio in a 2 x 2 table is

$$\hat{\theta} = \frac{n_{11} n_{22}}{n_{12} n_{21}} $$

For this reason, the odds ratio is also known as the _cross-product ratio_.


## Analysis: odds ratios


For our aspirin data, the odds of MI in the placebo group is about 83% higher than the odds of MI in the aspirin group:

```{r echo=TRUE}
(aspirin[1,1] * aspirin[2,2]) / 
  (aspirin[1,2] * aspirin[2,1])
```

Again, we'll use the `epitools` package for easier calculation of odds ratios and to obtain confidence intervals.    
\ 

Let's go to R!

## Example 2

Eight studies in China about smoking and lung cancer. (_Intern. J. Epidemiol._, 21: 197-201, 1992) We want to investigate the relationship between smoking and lung cancer while controlling for city.

```{r}
lung.cancer <- array(data = c(126,35,100,61,
                              908,497,688,807,
                              913,336,747,598,
                              235,58,172,121,
                              402,121,308,215,
                              182,72,156,98,
                              60,11,99,43,
                              104,21,89,36),
                     dim = c(2,2,8),
                     dimnames = list("smoking" = c("smokers","nonsmokers"),
                                  "lung.cancer" = c("yes","no"),
                                  "city" = c("Beijing","Shanghai","Shenyang",
                                             "Nanjing","Harbin","Zhengzhou",
                                             "Taiyuan","Nanchang")))
# Table 3.3
ftable(lung.cancer, row.vars = c("city","smoking"))
```

## Three-way contingency tables

The data on the previous slide have three categorical variables. The tabular layout is a _three-way contingency table_.    
\ 

This layout is useful when we want to compare two categorical variables while controlling for a third categorical variable.   
\ 

"Controlling for a third categorical variable" basically means holding it at a constant level while analyzing the other two categorical variables.    
\ 

The two-way table formed when controlling for a third variable is called a _partial table_.


## Some analysis options

1. Test the null hypothesis that smoking and lung cancer are conditionally independent given city, that is the odds ratio = 1 for each partial table (Cochran-Mantel-Haenszel Test)
2. Test the null hypothesis that the odds ratio between smoking and lung cancer is the same at each level of city. (Breslow-Day Test)
3. Estimate a _common odds ratio_ if association seems stable across the partial tables.

## Getting data into R

We demonstrate how to get this data into R in the R script that accompanies this workshop. The code doesn't fit comfortably onto slides.    
\ 


The manual method involves the `array` function that allows us to specify a third dimension, or third layer, of data.   
\ 

Once again we can use the `table` and `xtabs` functions to create our contingency tables if we import our data into a data frame.

## The `table` function for 3-way tables

```{r echo=FALSE}
set.seed(1)
dat3 <- data.frame(X = sample(c("Ctrl","Trt"),10, replace = T), Y = sample(c("Yes","No"),10, replace = T),
                  Z = sample(c("Male","Female"), 10, replace = T))
```

The first vector in the `table` function forms the rows, the second the columns, and each subsequent vector a layer.

```{r echo=TRUE}
table(dat3$X, dat3$Y, dat3$Z)
```

## The `xtabs` function for 3-way tables



```{r echo=TRUE}
# one row per obs
xtabs( ~ X + Y + Z, data = dat3) 
```

## The `xtabs` function for 3-way tables


```{r}
dat4 <- as.data.frame(table(dat3$X, dat3$Y, dat3$Z))
names(dat4)[1:3] <- c("X","Y","Z")
```

```{r echo=TRUE}
# one row per cell Freq
xtabs(Freq ~ X + Y + Z, data = dat4) 
```

## working with 3-way tables

Working with 3-way tables is a little trickier due to the third dimension. We can use the same functions, but we need to account for the third dimension.

- `prop.table` (cell proportions)
- `margin.table` (get table margins)
- `addmargins` (add table margins)
- `ftable` ("flatten" contingency tables into two dimensions)

## `prop.table` example

Specify `margin = c(1,3)` for row-wise proportions _within each two-way table_; specify `margin = c(2,3)` for column-wise proportions _within each two-way table_. Not specifying `margin` returns cell proportions summing to 1.

```{r echo=TRUE}
prop.table(lung.cancer, margin = c(1,3))
```

## `margin.table` example

Specify `margin = c(1,3)` for row-wise margins _within each two-way table_; specify `margin = c(2,3)` for column-wise margins _within each two-way table_. Not specifying `margin` sums all cell totals.

```{r echo=TRUE}
margin.table(lung.cancer, margin = c(1,3))

```

## `addmargins` example

The basic usage is to add row- and column-wise totals as well as the table total. Specifying `margin=1` or `margin=2` shows only row or column totals, respectively.

```{r echo=TRUE}
addmargins(lung.cancer)
```

## `ftable` example

By default, the third layer becomes the columns while the first and second layers are collapsed into the rows. You can specify which variables form the rows and columns using the `row.vars` and `col.vars` arguments.

```{r}
ftable(lung.cancer)
```



## Analysis: Cochran-Mantel-Haenszel (CMH) Test

Use the `mantelhaen.test` function to test the null hypothesis that smoking and lung cancer are conditionally independent given city (ie, odds ratio = 1 for each partial table)

```{r}
mantelhaen.test(lung.cancer)
```


## Common odds ratio

It is more informative to estimate the strength of association than test a hypothesis about it.   
\ 

The `mantelhaen.test` also returns an estimate of the _common odds ratio_ along with a 95% confidence interval. The common odds ratio is basically a weighted average of the partial table odds ratios.     
\ 

For the lung cancer data, the commons odds ratio is estimated as $2.17 (1.98, 2.38)$. The odds of lung cancer for smokers is about twice the odds of lung cancer for non-smokers.   
\ 

The CMH Test and common odds ratio estimate is really only appropriate when association is similar in each table.

## Analysis: Breslow-Day Test

We can use the `epi.2by2` function in the `epiR` package to test the null hypothesis that the odds ratio between smoking and lung cancer is the same at each level of city.    
\ 

As we'll see in the R script, the result of the Breslow-Day Test for our lung cancer data is not significant.    
\ 

This means we're justified in summarizing the conditional association by a single odds ratio for all eight partial tables.   
\ 

Let's go to R!

## Modeling categorical data

Another way to analyze categorical data is to _model_ it as a linear function of one or more explanatory variables.   
\ 

By _linear function_ we mean a weighted sum of variables. Say we model Y as a linear function of X and Z and obtain the following:   

$$Y = 0.65X + 0.04Z$$

This says we approximate Y by adding 0.65 * X and 0.04 * Z. 

## Logistic regression

The most common method of modeling categorical data is _logistic regression_.

- Binary logistic regression is for modeling binary outcomes (a categorical variable with two levels)
- Multinomial logistic regression is for modeling multiple unordered outcomes (a nominal categorical variable with more than two levels)
- Ordered logistic regression is for modeling multiple ordered outcomes (an ordered categorical variable with more than two levels)

We will just cover binary logistic regression.

## Example 3

Explore the relationship between age and presence/absence of coronary heart disease (CHD) in a study population.

```{r}
chd <- read.csv("chd.csv")
head(chd)
```

## Scatterplot of CHD by age

Difficult to describe this relationship.

```{r}
plot(chd ~ jitter(age), data=chd, ylim=c(0,1), axes = "F", xlab = "Age")
axis(side = 1, at = seq(20,70,10))
axis(side = 2, at = c(0,1))
box()
```

## Transform the data to better understand relationship

Calculate proportion of CHD within age groups.

```{r message=FALSE, cache=TRUE}
chd$agrp <- cut(x = chd$age, breaks = c(0, 29, 34, 39, 44, 49, 54, 59, 100), 
                labels = c("<29", "30-34","35-39","40-44","45-49","50-54","55-59",">60"))
library(dplyr)
chd2 <- chd %>% 
  group_by(agrp) %>% 
  summarise(n = n(), 
            absent = sum(chd==0), 
            present = sum(chd==1),
            proportion = round(present/n,2)) %>% 
  as.data.frame()
chd2
```


## Scatterplot of proportion with CHD by age group

```{r}
plot(seq(20,70, by=10),seq(0,1,by=0.2), type="n", xlab = "Age", ylab = "Proportion with CHD")
with(chd2, points(x = c(25, 32, 37, 42, 47, 52, 57, 65),
                  y = chd2$proportion))
```

## A function to model the relationship

Modeling the data in the previous graph essentially means finding a smooth mathematical function that describes the relationship between age and proportion with CHD.    
\ 

The function would allow us to estimate the probability of CHD in the study population given a certain age.   
\ 

The function would also allow us to summarize how the probability of CHD changes as age increases.    
\ 

The function needs to work such that plugging in age produces a number between 0 and 1. _This is basically what logistic regression ensures._    
\ 

## Performing logistic regression in R

We use the `glm` function with the `family` argument set to `binomial` to perform logistic regression in `R`.   
\ 

For the CHD, data this would be done as follows:

```{r echo=TRUE}
mod1 <- glm(chd ~ age, data = chd, family = binomial)
```

This performs binary logistic regression and saves the result to an object we called `mod1`.   
\ 

`chd ~ age` means "model chd as a function of age". 

## Viewing logistic regression results

Calling `summary` on our model object will return a summary of the logistic regression. 

```{r echo=TRUE, results='hide'}
summary(mod1)
```

The "formula" for our functional model is in the "Coefficients" section. For the model we fit:

$$P(\text{chd}) = -5.31 + 0.11 * \text{age}$$

Notice if we plug in an age = 60, we get chd = 1.29. That's because the response is on the _logit_, or _log-odds_, scale! 

## More on the logit

The logit transformation transforms values ranging from 0 to 1, to numbers ranging from $-\infty$ to $+\infty$.    
\ 

To make our model return a number between 0 and 1, we need to take the _inverse logit_. Fortunately R will do this for us. Here's how:

```{r echo=TRUE, results='hide'}
predict(mod1, type="response")
```

This will return the predicted probabilities for our data. We can also request predicted probabilities for certain age values.

```{r echo=TRUE, results='hide'}
predict(mod1, type="response", 
        newdata = data.frame(age = c(60,65,70)))
```

## Interpreting coefficients

If we exponentiate a variable's coefficient in a logistic regression model we get an odds ratio estimate of that variable's effect on the response.    
\ 

`exp(0.11) = 1.116`  
\ 

This says the odds of CHD increase by about 12% for every one year increase in age. To see how the odds change for every 10 years we could do this:    
\ 

`exp(0.11 * 10) = 3.004`    
\ 

The odds of CHD for a 50 year old are about 3 times that of a 40 year old. 

## Multiple logistic regression

The power of logistic regression is being able to include multiple independent variables in your model, categorical or continuous. This allows us to "control" for other variables.     
\ 

The odds ratio interpetation of the coefficients is the same, but now it is in the context of the additional variables being held constant.    
\ 

Simply add additional variables to your model with `+`. For example:   
\ 

`mod2 <- glm(chd ~ age + smoking + bp, data = chd,` 
`            family = binomial)`    
\ 

Let's go to R!


## Topics we haven't covered


## References

Agresti, A. _An Introduction to Categorical Data Analysis_. (1996) 

Hosmer, D. and Lemeshow, S. _Applied Logistic Regression_, 2nd Ed.. (2000)
